TSELM: Target Speaker Extraction using Discrete Tokens and Language Models

### 2024-12-23 src/paper.tex
u

There are only few works about discretization in Target Speech Extraction (TSE). SkiM-UniCATS \cite{gen_tse} is among the first to use discrete tokens in TSE, leveraging SSL models like HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, it overlooks the WavLM model, which excels in speech separation \cite{wavlm}. Also, it mainly focuses on single-layer outputs for discretization. Moreover, its evaluation is limited to speech quality, neglecting intelligibility and speaker similarity, both crucial for TSE. We try to address these potential issues by proposing TSELM, which uses multiple WavLM layers for discretization and adds metrics for intelligibility and speaker similarity. Our demos are available at 
\footnote{\href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}}.

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
We also compare our model with SkiM-UniCATS(vq-wav2vec) from \cite{gen_tse}, which uses discrete tokens from vq-wav2vec to conduct TSE. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech using our encoder and decoder. 
It serves as the upper bond for our model performance.  Besides TSELM, we conduct 3 main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-L6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross-attention without discretization. 
The concatenation strategy is still applied to the mixed speech. Mean Square Error loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but abandon the mixture concatenation strategy in the encoding step. The code and pretrained models are available at \footnote{\href{https://github.com/Beilong-Tang/TSELM}{https://github.com/Beilong-Tang/TSELM}}.

>>>>

There are only few works about discretization in Target Speech Extraction (TSE). SkiM-UniCATS \cite{gen_tse} is among the first to use discrete tokens in TSE, leveraging SSL models like HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, it overlooks the WavLM model, which excels in speech separation \cite{wavlm}. Also, it mainly focuses on single-layer outputs for discretization. Moreover, its evaluation is limited to speech quality, neglecting intelligibility and speaker similarity, both crucial for TSE. We try to address these potential issues by proposing TSELM, which uses multiple WavLM layers for discretization and adds metrics for intelligibility and speaker similarity. Our demos are available at 
\footnote{\href{https://tselmdemo.github.io/TSELM.demo/}{https://tselmdemo.github.io/TSELM.demo/}}.

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
We also compare our model with SkiM-UniCATS(vq-wav2vec) from \cite{gen_tse}, which uses discrete tokens from vq-wav2vec to conduct TSE. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech using our encoder and decoder. 
It serves as the upper bond for our model performance.  Besides TSELM, we conduct 3 main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-L6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross-attention without discretization. 
The concatenation strategy is still applied to the mixed speech. Mean Square Error loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but abandon the mixture concatenation strategy in the encoding step. The code and pretrained models are available at \footnote{\href{https://github.com/tselmdemo/tselm}{https://github.com/tselmdemo/tselm}}.
